---
title: "README.rmd"
author: "Phil Shea"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `binfunest`

<!-- badges: start -->
[![R-CMD-check](https://github.com/PhilShea/binfunest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PhilShea/binfunest/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PhilShea/binfunest/branch/main/graph/badge.svg)](https://app.codecov.io/gh/PhilShea/binfunest?branch=main)
<!-- badges: end -->

The goal of B2BQ is to simplify the estimation of offsets and Back-to-Back "Q" for communications systems.  The package also provides theoretical performance equations for many common modulation schemes, and eases maximum likelihood estimation of any function generating binomial probabilities.

## Installation

You can install the development version of B2BQ from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("PhilShea/binfunest")
```

## Example

This is a basic example which shows you how to solve a common problem of estimating the B2B Q and offset.  First we create a function which will add offset and B2B parameters to a standard communications modulation (Quadrature Phase Shift Keying, or QPSK).  The package provides theoretical performance curves for many common modulations.  Then `rbinom` is called to simulate measurements made at a range of signal to noise ratios (SNR).  All SNRs, offsets, and B2BQs are in Decibels.  Note that the generated samples include zeros, and that the first of these is useful in the estimate.

```{r example}
library(B2BQ, stats4)
QPSKdB.B2B <- B2BConvert( QPSKdB)
O1 <- 3 # offset
B1 <- 16 # B2BQ
s <- 0:20 # SNR Range 
N <- 1000000 # Number of samples
(r <- rbinom( length( s), N, QPSKdB.B2B( s, B1, O1)))
df <- data.frame( Errors=r, SNR=s, N=N) # place data in data frame

## This shows how you could do the work by hand
llsb2 <- function( b2b, offset)
       -sum( dbinom( r, N, QPSKdB.B2B( s, b2b, offset), log=TRUE))
mle1 <- stats4::mle( llsb2, start=c( b2b=20, offset=0), nobs=length(s),
                   method="Nelder-Mead")
stats4::coef( mle1)
# Below is the new function
est1 <-  mleB2B( data=df, Errors="Errors", N=N, f=QPSKdB.B2B,
                 fparms=list( x="SNR"), start=c(b2b=20, offset=0))

(est1coef <- stats4::coef( est1))
```

The plot below compares the theoretical curve to the curve with B2B and offset, and the estimated curve.


```{r plotQab}
plot( s, y=r/N, log='y', type='p', panel.first = grid())
lines( s, QPSKdB( s))
lines( s, QPSKdB.B2B( s, B1, O1), col='red')
lines( s, y=QPSKdB.B2B( s, est1coef[1],  est1coef[2]), col="green")
legend( "bottomleft",
        legend=c( "Data",  "Theory", "3 dB Offset + 16 dB B2B", 
                  "Estimated"),
        lty=c( NA, 1, 1, 1), col=c( 'black', 'black', 'red', 'green'),
        pch=c( 1, NA, NA, NA))
```
