---
title: "binfunest Release"
author: "Phil Shea"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require( binfunest)
set.seed( 31394)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `Introduction`

<!-- badges: start -->
[![R-CMD-check](https://github.com/PhilShea/binfunest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PhilShea/binfunest/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PhilShea/binfunest/branch/main/graph/badge.svg)](https://app.codecov.io/gh/PhilShea/binfunest?branch=main)
[![CRAN status](https://www.r-pkg.org/badges/version/binfunest)](https://CRAN.R-project.org/package=binfunest)
<!-- badges: end -->

The goal of `binfunest` is to simplify the estimation of offsets and Back-to-Back "Q" for communications systems.  The package also provides theoretical performance equations for many common modulation schemes, and eases maximum likelihood estimation of any function generating binomial probabilities.

## Installation

`binfunest` is available on CRAN, so `install.packages( "binfunest")` will install it, and `library( binfunest)` will make it available. You can install the development version of `binfunest` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("PhilShea/binfunest")
```

## Example

This is a basic example which shows how to solve a common problem of estimating the B2B Q and offset.  First we create a function which will add offset and B2B parameters to a standard communications modulation (Quadrature Phase Shift Keying, or QPSK).  The package provides theoretical performance curves for many common modulations.  Then `rbinom` is called to simulate measurements made at a range of signal to noise ratios (SNR).  All SNRs, offsets, and B2BQs are in Decibels.  Note that the generated samples include zeros, and that the first of these (i.e., the lowest SNR that produced zero errors) is useful in the parameter estimate.

```{r example}
require( binfunest)
require( stats4)
QPSKdB.B2B <- B2BConvert( QPSKdB)
O1 <- 3 # offset
B1 <- 15 # B2BQ
s <- 0:25 # SNR Range 
N <- 1000000 # Number of samples
(r <- rbinom( length( s), N, QPSKdB.B2B( s, B1, O1)))
df <- data.frame( Errors=r, SNR=s, N=N) # place data in data frame

## This shows how you could do the work by hand
llsb2 <- function( b2b, offset)
       -sum( dbinom( r, N, QPSKdB.B2B( s, b2b, offset), log=TRUE))
mle1 <- stats4::mle( llsb2, start=c( b2b=20, offset=0), nobs=length(s),
                   method="Nelder-Mead")
stats4::coef( mle1)


# Below is the new function
est1 <-  mleB2B( data=df, Errors="Errors", N=N, f=QPSKdB.B2B,
                 fparms=list( x="SNR"), start=c(b2b=20, offset=0))

(est1coef <- stats4::coef( est1))
```

The plot below compares the theoretical curve to the curve with the chosen B2B and offset, and the curve with the estimated parameters. Note that all samples above 18 dB SNR returned zero, except the 21 dB point.  The Estimated line is dashed so the 3 dB + Offset line, which it overlays, is visible.

```{r plotQab, echo=FALSE, message=FALSE, warning=FALSE}
plot( s, y=r/N, log='y', type='p', panel.first = grid(), ylim=c(1e-9,0.5),
      main="Modulation Performance Curves", xlab="SNR in dB",
      ylab="BER")
lines( s, QPSKdB( s))
lines( s, QPSKdB.B2B( s, B1, O1), col='red')
lines( s, QPSKdB.B2B( s, +Inf, O1), col='blue')
lines( s, y=QPSKdB.B2B( s, est1coef[1],  est1coef[2]), col="green", lty=2)
abline( h=QPSKdB.B2B( B1, +Inf, O1), col='black', lty=2)
legend( "bottomleft",
        legend=c( "Data",  "Theory", "3 dB Offset", "3 dB Offset + 15 dB B2B", 
                  "Estimated", "B2B Limit"),
        lty=c( NA, 1, 1, 1, 2, 2), col=c( 'black', 'black', 'blue', 'red',
                                          'green'),
        pch=c( 1, NA, NA, NA, NA, NA))
```
